Analysis of professional surfing tournaments with generalized Bradley-Terry

### 2021-08-12 inst/surfing_jsa.tex
more work on details

@@

\cite{farley2013} analyse surfing results using one-way ANOVA techniques and
report comparisons between the top 10 and bottom 10 surfers in the
2013 World Championship Tour and conclude that the top-ranked athletes
are more consistent than the lower-ranked group. They report that "To
date, only a limited number of research studies have reported on the
results of competitive surfing" [p44], and the analysis presented here
is hoped to change that.

Other statistical research into competitive surfing appears to focus
on the scoring of aerial manoeuvres compared with other manoeuvres in
events.  \cite{undgren2014}, for example, report that aerial manoeuvres
scored higher than other stunts but had a lower completion rate, a
theme we return to below.

@@

Table~\ref{resultstable} shows the official results table from the
tour.  We see that Ferreira places first with 59740 points, Medina
second with 56475, and so on.  The overall ranking of the surfers is
determined by summing the season's points.  However, this system
suffers from the defect that the points system itself is intrinsically
arbitrary: the details of the points system does not affect the
competitors' behaviour but can change the overall ranking.
Plackett-Luce likelihood functions are not subject to such a
phenomenon and furnish an objective and coherent method for ranking
competitors.  Further, likelihood-based methods offer a further
advantage over points-based systems: the ability to conduct
statistically rigorous tests of meaningful nulls.

@@

\begin{equation}
\frac{
p_\mathrm{Andino}^{19}\, p_\mathrm{Callinan}^{19}\, p_\mathrm{Coffin}^{8}\ldots
}{
(p_\mathrm{Andino} + p_\mathrm{Bourez})  (p_\mathrm{Andino} + p_\mathrm{Buchan})(p_\mathrm{Bourez} + p_\mathrm{Flores} + p_\mathrm{Freestone})\ldots
}
\end{equation}

\begin{figure}
\includegraphics[width=4in]{surfing_files/figure-html/showsurfingmaxppie-3.png} % Pie chart of strengths
\caption{Pie chart\label{piechartmax} showing maximum likelihood strengths of the 23 competitors.  Note the
dominance of Ferreira, Florence, and Medina}
\end{figure}

Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
hyper2} software may be used to formally assess one plausible null:
that the competitors all have equal skill, that is $H_0\colon
p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  The Method of Support \cite{edwards1973}
rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
There is thus strong evidence to suggest that the competitors are
indeed of different skills and the result is not random.  Note that
the competitors' points (that is, the final column of
table~\ref{resultstable}) does not admit this type of analysis, as the
points awarded are arbitrary.

@@

The statistical method described here can test many different
suggestions drawn from surfing lore {\bf Tim is there a better way to
  describe this?}.  For example, one frequent suggestion {\bf Tim I
  need citations here} is that Brazilian surfers such as Ferreira and
Toledo have surfing skillsets that favour point and beach break
beaches, those types of beach being readily accessible to these
surfers' native Brazil.  We now assess the null that Brazilians
perform equally well on all types of beaches.  To that end, we create
a reified surfer, ``Brazilian beachtype", whose strength helps all
Brazilian surfers when surfing on their (putative) preferred
beachtype, analogously to the ``home ground monster" who helps the
home team as per equation~\ref{home_ground_advantage}.  The null would
be that this reified entity has zero strength.  We fail to reject this
null (differential support = 0.035; $p=0.79$) and conclude that there
is no evidence to suggest that Brazilians perform is better on their
preferred beach types than one would expect by chance.  We may go
further and state that not only is the maximum likelihood estimate for
Brazilian beachtype is very small (about 0.002), but we may rule out
the suggestion that the strength of this entity exceeds about 0.025
(Figure~\ref{brazilian_beachtype_support}).

\begin{figure}
  \includegraphics[width=4in]{surfing_files/figure-html/plotwidep-1.png}
\caption{Support \label{brazilian_beachtype_support} for the Brazilian beachtype reified entity}
\end{figure}

@@

>>>>

\section*{Abstract}

{\bf\large Tim we need to discuss this}

@@

Farley {\em et. al.} \cite{farley2013} analyse surfing results using
one-way ANOVA techniques and report comparisons between the top 10 and
bottom 10 surfers in the 2013 World Championship Tour and conclude
that the top-ranked athletes are more consistent than the lower-ranked
group.  They report that ``to date, only a limited number of research
studies have reported on the results of competitive surfing" [p44],
and the analysis presented here fills some of this gap.

Other statistical research into competitive surfing appears to focus
on the scoring of aerial manoeuvres compared with other manoeuvres in
events.  Lundgren {\em et. al.} \cite{lundgren2014}, for example,
report that aerial manoeuvres scored higher than other stunts but had
a lower completion rate, a theme we return to below.

@@

Table~\ref{resultstable} shows the official results table from the
tour.  We see that Ferreira places first with 59740 points, Medina
second with 56475, and so on.  The overall ranking of the surfers is
determined by summing the season's points.  However, this system
suffers from the defect that the points system itself is intrinsically
arbitrary: the details of the points system does not affect the
competitors' behaviour but can change the overall ranking; this would
suggest that more formal statistical methods are neede for analysis of
such datasets.  The Plackett-Luce likelihood functions are not subject
to such a phenomenon and furnish an objective and coherent method for
ranking competitors.  Further, likelihood-based methods offer one
advantage over points-based systems: the ability to conduct
statistically rigorous tests of meaningful nulls.

@@

\begin{equation}\label{complete_likelihood_function}
\frac{
p_\mathrm{Andino}^{19}\, p_\mathrm{Callinan}^{19}\, p_\mathrm{Coffin}^{8}\ldots
}{
(p_\mathrm{Andino} + p_\mathrm{Bourez})  (p_\mathrm{Andino} + p_\mathrm{Buchan})(p_\mathrm{Bourez} + p_\mathrm{Flores} + p_\mathrm{Freestone})\ldots
}
\end{equation}

\subsection{Bradley-Terry and pairs}

Paired comparisons are particularly favourable to Bradley-Terry
because the likelihood function is exponential family and as such is
amenable to analysis via general linear models~\cite{turner2012}.
However, this breaks down when more than two entities are compared and
here {\bf clumsy wording} we see comparisons between three surfers
and, in the case of wct08, many more.  It is natural to ask how much
information is contained in the observations where more than two
surfers are in the water.

We can directly calculate the determinant of the observed information
matrix for the likelihood function in
equation~\ref{complete_likelihood_function} at about $4.3\times
10^{71}$, as compared with~$2.9\times 10^{69}$ for the pairs
considered in isolation, a ratio of about $146.4$.  We are thus
motivated to consider the whole dataset, not just the part which
involves paired comparisons.  {\bf clumsy wording}.

\section{Results}
\begin{figure}
\includegraphics[width=4in]{surfing_files/figure-html/showsurfingmaxppie-3.png} % Pie chart of strengths
\caption{Pie chart\label{piechartmax} showing maximum likelihood strengths of the 23 competitors.  Note the
dominance of Ferreira, Florence, and Medina}
\end{figure}

Figure~\ref{piechartmax} shows the maximum likelihood estimate for the
23 competitors' strengths, which appears to show a wide range from
Florence at about 17.5\% down to Coffin at about 1.5\%.  The {\tt
  hyper2} software may be used to formally assess one plausible null:
that the competitors all have equal skill, that is $H_0\colon
p_\mathrm{Andino} = p_\mathrm{Bourez}=\ldots=
p_\mathrm{Wright}=\frac{1}{23}$, and any differences in placings are
due to random variation.  The Method of Support \cite{edwards1992}
rejects the null (in favour of the evaluate) at $p=4\times 10^{-5}$.
There is thus strong evidence to suggest that the competitors are
indeed of different skills and the result actually reflects genuine
differences in abilities, rather than the result of random chance.
Observe that the competitors' points (that is, the final column of
table~\ref{resultstable}) cannot be used for this type of
probabalistic analysis: the points awarded are arbitrary and changing
the points system would change the points scored but not affect the
competitors' behaviour.

@@

The statistical method described here can test many different
suggestions drawn from surfing lore {\bf Tim is there a better way to
  describe this?}.  For example, one frequent suggestion {\bf Tim I
  need citations here} is that Brazilian surfers such as Ferreira and
Toledo have surfing skillsets that favour point and beach break
beaches {\bf Tim I need some help describing and defining these here},
those types of beach being readily accessible to these surfers' native
Brazil.  We now assess the null that Brazilians perform equally well
on all types of beaches.  To that end, we create a reified surfer,
``Brazilian beachtype", whose strength helps all Brazilian surfers
when surfing on their (putative) preferred beachtype, analogously to
the ``home ground monster" who helps the home team as per
equation~\ref{home_ground_advantage}.  The null would be that this
reified entity has zero strength.  We fail to reject this null
(differential support = 0.035; $p=0.79$) and conclude that there is no
evidence to suggest that Brazilians' performce is better on their
preferred beach types than one would expect by chance.  We may go
further using the method of support and state that not only is the
maximum likelihood estimate for Brazilian beachtype is very small
(about 0.002), but we may rule out the suggestion that the strength of
this entity exceeds about 0.025 (Figure~\ref{brazilian_beachtype_support}).

\begin{figure}
  \includegraphics[width=4in]{surfing_files/figure-html/plotwidep-1.png}
\caption{Profile support \label{brazilian_beachtype_support} for the
  Brazilian beachtype reified entity.  The horizontal axis specifies a
  particular strength for Brazilian beachtype, and the height of the
  curve gives the support conditional on the x-axis}
\end{figure}

@@

{\bf\large Tim I am on thin ice here, I need to talk with you and
  clarify this}

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics, and it is natural to ask whether this would
affect the competition in some way.  One suggestion {\bf Tim?} would
be that competitors would become more focussed on beating compatriots
than surfers from other countries, the thinking being that admission
to a country's Olympic squad would be of more importance to a
competitor than their result in wct.  {\bf Tim this is clumsy
  wording}.  We would expect competitors to value their ranking when
restricted to within their country more than their ranking over all of
wct.  If this is the case then we would expect competitors to change
their behaviour if they are in the water with another surfer of the
same nationality.  Specifically, if we have three surfers in the water
of whom two are of one nationality and the third of another, we would
expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.  {\bf Tim this is poor wording we need to rephrase}.

Such hypotheses are difficult to test with conventional techniques as
competitors's strengths are nuisance parameters ~\cite{basu1977}.
However, the

Consider, as an example, the case of Smith, Buchan, Freestone.  Buchan
and Freestone are Australian, and Jordy Smith is South African.  It is
reasonable to ask whether Buchan and Freestone, who are in direct
competition with one another for a place in the Australian Olympic
surfing squad, would be more motivated to beating each other than to
come ahead of Smith.  This would be a rational response, but arguably
noncompetitive.  How would such a strategy manifest itself in
scorelines?

Reified Bradley-Terry technique furnishes an intuitive statistical
test for such a mechanism which controls for competitors' strengths.
Noting that the compatriots effectively place themselves at a {\em
  disadvantage} when compared with the odd man out, we introduce a
compatriot monster whose strength $p_C$ helps the odd man out.
Writing $p_1, p_2$ for the strengths of the compatriots and $q$ for
the odd man out, we construct likelihood functions for the six
possible orderings when two compatriots and an odd man out are in the
water at the same time in Table~\ref{likelihoodcompatriot} (if we have
either three different nationalities or three identical nationalities,
we revert to the standard Plackett-Luce likelihood).  Note that all
six likelihood functions in table~\ref{likelihoodcompatriot} are
informative about $p_1,p_2,q$ in addition to being informative about
the object of inference, $p_C$.

\begin{table}
\begin{tabular}{l|l}
order  & ${\mathcal L}(p_1,p_2,q,p_C)$\\ \hline
$p_1\succ p_2\succ q $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{p_2}{p_2+q+p_C}$\\
$p_2\succ p_1\succ q $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{p_1}{p_1+q+p_C}$\\
$p_1\succ q\succ p_2 $ & $ \frac{p_1}{p_1+p_2+q+p_C}\cdot\frac{q+p_C}{p_2+q+p_C}$\\
$p_2\succ q\succ p_1 $ & $ \frac{p_2}{p_1+p_2+q+p_C}\cdot\frac{q+p_C}{p_1+q+p_C}$\\
$q\succ p_1\succ p_2 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_1}{p_1+p_2}$\\
$q\succ p_2\succ p_1 $ & $ \frac{q+p_C}{p_1+p_2+q+p_C}\cdot\frac{p_2}{p_1+p_2}$\\
\end{tabular}
\caption{The \label{likelihoodcompatriot} six possible finishing
  orders (``$\succ$'' means ``scored more points than'') for the case
  where two compatriot surfers and an odd man out are in the water at
  the same time, with associated likelihood function for strengths
  $p_1,p_2,q, p_C$.  Here $p_1,p_2$ are the strengths of two
  compatriot surfers, $q$ the strength of the odd man out, $p_C$ the
  strength of the compatriot monster. and nationality and a third are
  in the water at the same time}
\end{table}

We can now apply the Method of Support to the hypothesis that $p_C=0$,
that is, that this noncompetitive behaviour is absent.  We find that
the maximum likelihood estimate for $p_C$ is about 0.015, weaker than
the ``real'' surfers, but the support for a nonzero $p_C$ is about 1.8
[asymptotic $p=0.28$], short of the two units of support required for
rejection of our null.  We conclude that there is no strong evidence
to suggest that such noncompetitive behaviour occurred, at least not
from the order statistics we considered.

\section{Conclusions and further work}

{\bf\large Tim we need to discuss this}
